Joint estimation of phylogenies and ancestral states is not consistent

### 2017-12-14 appendix.tex
address (almost all) comments

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation.
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
We start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.
%EM is there a logic to what gets bolded or no? siteSplit isn't, but a site pattern is.

@@

For the $i$th factor of \eqref{eq:full_likelihood},
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
%EM continuing to nitpick, is there anything that is just s_j? Seems funny to only have the tilde s_j form.
\begin{align*}
    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$ and
\begin{align*}
    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this we have
\begin{align*}
\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

@@

Let
$$
%EM It seems a little strange that eta appears here as an argument when it didn't appear in eq:site_pattern_likelihood. In fact it's not really an argument, right? Everything is determined by t and tau it seems to me.
L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t)) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t)
$$
and the partial log-likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
\end{equation}

@@

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
%EM Nitpick-- eta takes a tau and a t as arguments. Here we give it a p.
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, p)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, p)$ we have
$$
P(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_5=\{1,2\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^3(1-p)^2.
$$
Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, p)$,
$$
P(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3.
$$

@@

%EM I think this following statement could use a little more justification. If we write out a symmetric function f in two variables and take the derivative of f(x+delta, x) = f(x, x+delta) with respect to delta and set it to zero we can see that this must be a stationary point of f. Do we need to say more to ensure that this is a maximum?
Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
The analogous statement holds for $y_1$ and $y_2$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

We begin by writing out the case for the ancestral state split $\{1\}$.

@@

Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_likelihood}, the log-likelihood is

@@

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
%EM It would seem helful for the below check to express these a_1' etc in terms of tilde{p}.
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}. \\
    \end{aligned}
    \label{eq:a_const_prime}
\end{equation}
%EM This check would make a lot more sense if you put the C^{2}_{\tau_1,\tau_1}(x^*, y^*)  and the other analogous definition first.
%EM It would also be nice to do one of the calculations to get us from a_1 world to a_1' world. Just staring at table S4 doesn't make this obvious.
We also now need to check that
$$
\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
$$
are both between zero and one.
It will suffice to show
$$
0 \le \tilde{p} - 2p_{13} \le 1
$$
and
$$
0 \le \tilde{p} + 2p_{13} \le 1
$$
for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
Using Table~\ref{tab:sitepatprob} and $t^*$,
$$
p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right).
$$
We see that
\begin{equation}
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 2 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - (x^*)^2.
\label{eq:lenny}
\end{equation}
Note
$$
(y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - 4x^* + (x^*)^2;
$$
showing that \eqref{eq:lenny} holds if $x^* \le 2$, which is satisfied by definition.
Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}$ and $\hat{y}$ are valid.

@@

We observe that

@@

always, and thus the maximum can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.

@@

>>>>

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation.
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
We start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.
%EM is there a logic to what gets bolded or no? siteSplit isn't, but a site pattern is.
%das: I was thinking bold for matrices/vectors (here of character data) then a vector of character data maps to a set (site split), and so I didn't bold the set.

@@

For the $i$th factor of \eqref{eq:full_likelihood},
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
%EM continuing to nitpick, is there anything that is just s_j? Seems funny to only have the tilde s_j form.
%das: I think s_j used to be the site pattern and then \tilde{s}_j became the split? It's still the case for h (no tilde is ancestral state pattern, tilde is ancestral state split). I changed to \tilde{y} to be consistent with this, but mabye we can think of something better?
\begin{align*}
    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$ and
\begin{align*}
    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this we have
\begin{align*}
\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

@@

Let
$$
%EM It seems a little strange that eta appears here as an argument when it didn't appear in eq:site_pattern_likelihood. In fact it's not really an argument, right? Everything is determined by t and tau it seems to me.
%das: We added it to clarify that the likelihood now depends on \xi, which is an element of \eta. It seems dirty to ignore it. At the risk of adding more notation, we could add \hat{H} to the original likelihood? Or just axe the etas...
L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t)) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t)
$$
and the partial log-likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
\end{equation}

@@

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch, with $t=\{p,p,p,p,p\}$.
The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
%EM Nitpick-- eta takes a tau and a t as arguments. Here we give it a p.
%das: nit unpicked
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, t)$ we have
$$
P(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_5=\{1,2\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^3(1-p)^2.
$$
Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, t)$,
$$
P(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3.
$$

@@

%EM I think this following statement could use a little more justification. If we write out a symmetric function f in two variables and take the derivative of f(x+delta, x) = f(x, x+delta) with respect to delta and set it to zero we can see that this must be a stationary point of f. Do we need to say more to ensure that this is a maximum?
%das: I wasn't even thinking derivatives. Let f(x, y) be symmetric so that f(x, y) = f(y, x). If f is maximized at some (x_0, y_0) then f(x_0, y_0) > f(x, y) for all x, y. But then by symmetry f(y_0, x_0) is also a maximum. Now *if* a unique maximum exists, x_0 = y_0 should necessarily be this maximum by the properties defined above, otherwise, f(x_0, y_0) would not be equal to f(y_0, x_0). I think there are probably counterexamples where there is not a unique maximum where this statement isn't necessarily true (trivially, because you'd just say x_0 = y_1 if y_1 is also a maximum and voila) but I don't think this discounts the fact that x_0 = y_0 will still be *a* maximum in this case. In that case, it's reasonable to assume so, in my opinion, especially for the topological inconsistency (we don't use the branch length values), but even for the branch length inconsistency because the case of no unique solution is just as bad if not worse than inconsistency.
%das: Should some or all of that be in the text? Can you see anything wrong with it?
Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
The analogous statement holds for $y_1$ and $y_2$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

We begin by writing out the case for the ancestral state split $\{1\}$, and show the other two cases follow a similar argument.

@@

%das: gah, you're right
Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_profile_likelihood_mean}, the log-likelihood is

@@

%das: done

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
%EM It would seem helful for the below check to express these a_1' etc in terms of tilde{p}.
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}. \\
    \end{aligned}
    \label{eq:a_const_prime}
\end{equation}
%EM This check would make a lot more sense if you put the C^{2}_{\tau_1,\tau_1}(x^*, y^*)  and the other analogous definition first.
%EM It would also be nice to do one of the calculations to get us from a_1 world to a_1' world. Just staring at table S4 doesn't make this obvious.
%das:todo
We also now need to check that
$$
\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
$$
are both between zero and one.
It will suffice to show
$$
0 \le \tilde{p} - 2p_{13} \le 1
$$
and
$$
0 \le \tilde{p} + 2p_{13} \le 1
$$
for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
Using Table~\ref{tab:sitepatprob} and $t^*$,
$$
p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right).
$$
We see that
\begin{equation}
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 2 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - (x^*)^2.
\label{eq:lenny}
\end{equation}
Note
$$
(y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - 4x^* + (x^*)^2;
$$
showing that \eqref{eq:lenny} holds if $x^* \le 2$, which is satisfied by definition.
Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}$ and $\hat{y}$ are valid.

@@

Replacing all $1+w$ with $1-w$ in Table~\ref{tab:likelihoods} for $\tau = \tau_2$ will result in a lower bound on the maximum, so we proceed with that first.
Then, for site split $\{1,2\}$ we have
\max\left((1+x)^2 (1-w)(1+y)^2, (1+x)(1-x)(1-w)(1+y)(1-y)\right)
where since

@@

%das: Okay, think I missed the point about 1+w for 1-w that I added back.
always, the maximum can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.

@@

%das: This is old and I believe the text has been updated to fix it. Can you verify?
